---
title: "extendedCosinorVignette"
output: rmarkdown::html_vignette
vignette: >
    %\VignetteIndexEntry{extendedCosinorVignette}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---
    
```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.height = 6,
    fig.width = 10
)
```

```{r setup}
library(extendedCosinor)
library(tidyverse)
library(cluster)
```

## Data Requirements

We'll simulate seven days of minute-level data `x` for one person for this demonstration. 
`cosinorExtendedModel` and `cosinorBasicModel` both take a vector of epoch-level data as input, 
as well as the argument `window` which specifies the epoch duration in minutes.
The vector must have a length that is an integer multiple of `1440/window`.
The model assumes the first element of x corresponds to the first epoch after midnight.

```{r}
t_hours <- seq(0, 24*7,by = 1/60)[-1]
x <- sqrt(exp(2 + 3*cos((t_hours- 5)/24*2*pi) + rnorm(length(t_hours),1)))
plot(x~t_hours)
```

```{r}
cosOut <- cosinorExtendedModel(
    x = x, 
    window = 1, 
    export_ts = TRUE,
    export_cosinor_param = TRUE,
    export_nls_outcome = TRUE
)
cosOut$estimates
```

```{r}
cosOut$cosinor_ts
```

## Interpreting the Output

### Primary Model Parameter Estimates

The five primary model parameter estimates are the following:

```{r}
cosOut$estimates %>% select(mu_ext:phi_ext)
```

These correspond to the model outlined by Marler et al (2006). 
We've renamed some of the parameters ($\mu_ext$ was "min" and $\gamma_ext$ was "amp") from the original version to avoid confusion
with the amplitude and minimum of the fitted function.

$$
    y(t) = \mu_{\mathrm{ext}} + \gamma_{\mathrm{ext}}\cdot \mathrm{expit}\bigg(\beta_{\mathrm{ext}} \Big(\cos(\left[t - \phi_{\mathrm{ext}}\right]\cdot 2\pi/24)-\alpha_{\mathrm{ext}} \Big) \bigg) + \epsilon(t)
$$
    
The plot below shows that $mu_ext$ (formerly "min") can be close to the function minumum,
but it is not exact.
Additionally, $gam_ext$ (formerly "amp") influences the function amplitude,
but it is not the amplitude of the fitted function.

```{r, warning = FALSE, message = FALSE}
cosOut$cosinor_ts %>% filter(
    time_across_days <= 24
) %>% ggplot() + geom_point(
    aes(
        x = time_across_days, y = y
    ), alpha = 0.2, color = "blue"
)  + geom_line(
    aes(
        x = time_across_days, y = y_ext
    )
) + scale_x_continuous(
    "Time of Day (hours since midnight)",
    breaks = 4*(0:6)
) + geom_hline(
    "hline",
    yintercept = c(
        cosOut$estimates$mu_ext
    ), color = "red"
) + geom_segment(
    aes(
        xend = cosOut$estimates$acrotime_ext+12,
        x = cosOut$estimates$acrotime_ext+12,
        y = cosOut$estimates$mu_ext,
        yend = cosOut$estimates$mu_ext + 
            cosOut$estimates$gamma_ext
    ), color = "purple", arrow = arrow(ends = "both")
) + geom_label(
    data = tibble(
        name = c("mu_ext", "mu_ext + gamma_ext"),
        x = c(cosOut$estimates$acrotime_ext, cosOut$estimates$acrotime_ext + 12),
        y = c(
            cosOut$estimates$mu_ext, 
            cosOut$estimates$mu_ext + cosOut$estimates$gamma_ext + 3
        )
    ),
    aes(
        x = x,
        y = y,
        label = name
    )
) + ggtitle(
    "Primary Estimates from extendedCosinor"
)
```

### Secondary Estimates

Secondary estimates are more appropriate for analysis, and are derived from 
the fitted function and residuals.

```{r}
cosOut$estimates %>% select(acrotime_ext:up_mesor_deriv_ext,F_pseudo_ext,R2_ext)
```

The connection of each estimate to the fitted function is depicted below.

```{r, warning = FALSE, message = FALSE}
label.df = cosOut$estimates %>% select(
    acrotime_ext:up_mesor_deriv_ext
) %>% pivot_longer(
    acrotime_ext:up_mesor_deriv_ext
) %>% mutate(
    x = ifelse(
        name %in% c(
            "minimum_ext",
            "mesor_ext",
            "maximum_ext"
        ), 20, value
    ), 
    y = ifelse(
        !name %in% c(
            "minimum_ext",
            "mesor_ext",
            "maximum_ext"
        ), 60 + c(0,-15,3, 6), value
    ),
    y = ifelse(
        name == "amplitude_ext",
        cosOut$estimates$maximum_ext*0.75,
        y
    ),
    x = ifelse(
        name == "amplitude_ext",
        cosOut$estimates$acrotime_ext+11.5,
        x
    ),
    y = ifelse(
        name == "hours_above_mesor_ext",
        38,
        y
    ),
    x = ifelse(
        name == "hours_above_mesor_ext",
        cosOut$estimates$acrotime_ext,
        x
    ),
    y = ifelse(
        name == "up_mesor_deriv_ext",
        cosOut$estimates$mesor_ext + (15*cosOut$estimates$up_mesor_deriv_ext),
        y
    ),
    x = ifelse(
        name == "up_mesor_deriv_ext",
        cosOut$estimates$up_mesor_ext + 15,
        x
    )
)

cosOut$cosinor_ts %>% filter(
    time_across_days <= 24
) %>% ggplot() + geom_point(
    aes(
        x = time_across_days, y = y
    ), alpha = 0.2, color = "blue"
)  + geom_line(
    aes(
        x = time_across_days, y = y_ext
    )
) + scale_x_continuous(
    "Time of Day (hours since midnight)",
    breaks = 4*(0:6)
) + geom_hline(
    "hline",
    yintercept = c(
        cosOut$estimates$minimum_ext,
        cosOut$estimates$mesor_ext,
        cosOut$estimates$maximum_ext
    ), color = "red"
) + geom_vline(
    xintercept = c(
        cosOut$estimates$up_mesor_ext,
        cosOut$estimates$down_mesor_ext,
        cosOut$estimates$acrotime_ext
    ), color = "orange"
) + geom_segment(
    aes(
        xend = cosOut$estimates$down_mesor_ext,
        x = cosOut$estimates$up_mesor_ext,
        y = 40, yend = 40
    ), color = "green", arrow = arrow(ends = "both")
) + geom_segment(
    aes(
        xend = cosOut$estimates$acrotime_ext+12,
        x = cosOut$estimates$acrotime_ext+12,
        y = cosOut$estimates$minimum_ext,
        yend = cosOut$estimates$minimum_ext +
            cosOut$estimates$amplitude_ext
    ), color = "purple", arrow = arrow(ends = "both")
) + geom_abline(
    aes(
        intercept = cosOut$estimates$mesor_ext - 
            cosOut$estimates$up_mesor_deriv_ext*
            cosOut$estimates$up_mesor_ext,
        slope = cosOut$estimates$up_mesor_deriv_ext
    ), color = "cyan"
) + geom_label(
    data = label.df,
    aes(
        x = x,
        y = y,
        label = name
    )
) + ggtitle(
    "Secondary Estimates from extendedCosinor"
)
```



### Comparison to ActCR::ActExtendCosinor

```{r}
actcrOut <- ActCR::ActExtendCosinor(
    x = x, 
    window = 1, 
    export_ts = TRUE
)
actcrOut$params %>% as.data.frame
```

The plot below shows how certain estimates from the `ActCR::ActExtendCosinor` output map to 
the fitted function. 
The plot below confirms that the estimates labelled `amp` and `minimum` do not correspond
to the amplitude and minimum of the fitted function. 
Additionally, the estimate of `MESOR`, `UpMesor`, and `DownMesor` do not seem to capture
the intended construct--the midline of rhythm.

```{r, warning = FALSE, message = FALSE}
actcrOut$cosinor_ts %>% filter(
    time_across_days <= 24
) %>% ggplot() + geom_point(
    aes(
        x = time_across_days, y = original
    ), alpha = 0.2, color = "blue"
)  + geom_line(
    aes(
        x = time_across_days, y = fittedYext
    )
) + scale_x_continuous(
    "Time of Day (hours since midnight)",
    breaks = 4*(0:6)
) + geom_hline(
    "hline",
    yintercept = c(
        actcrOut$params$minimum,
        actcrOut$params$MESOR,
        actcrOut$params$minimum + actcrOut$params$amp
    ), color = "red"
) + geom_vline(
    xintercept = c(
        actcrOut$params$acrotime,
        actcrOut$params$UpMesor,
        actcrOut$params$DownMesor
    ), color = "orange"
) + geom_segment(
    aes(
        xend = actcrOut$params$acrotime + 12,
        x = actcrOut$params$acrotime + 12,
        y = actcrOut$params$minimum,
        yend = actcrOut$params$minimum + actcrOut$params$amp
    ), color = "purple", arrow = arrow(ends = "both")
)  + geom_label(
    data = tibble(
        name = c("minimum", "amp", "acrotime", "MESOR", "UpMesor", "DownMesor"),
        x = c(
            20, actcrOut$params$acrotime + 12, actcrOut$params$acrotime, 
            20, actcrOut$params$UpMesor, actcrOut$params$DownMesor
        ),
        y = c(
            actcrOut$params$minimum, 
            actcrOut$params$minimum + actcrOut$params$amp + 3,
            actcrOut$params$MESOR*2 + 15,
            actcrOut$params$MESOR,
            actcrOut$params$MESOR*2+10,
            actcrOut$params$MESOR*2 + 5
        )
    ),
    aes(
        x = x,
        y = y,
        label = name
    )
) + ggtitle(
    "Estimates from ActCR::ActExtendCosinor"
)
```

### Cosinor Model Estimates

By setting the option `export_cosinor_param = TRUE`, the function includes basic cosinor parameter estimates
in the output. These estimates are mapped onto the basic cosinor fit below.

```{r, warning = FALSE, message = FALSE}
cosOut$cosinor_ts %>% filter(
    time_across_days <= 24
) %>% ggplot() + geom_point(
    aes(
        x = time_across_days, y = y
    ), alpha = 0.2, color = "blue"
)  + geom_line(
    aes(
        x = time_across_days, y = y_ext
    ), linetype = 2
) + geom_line(
    aes(
        x = time_across_days, y = y_cos
    ), linewidth = 1
) + scale_x_continuous(
    "Time of Day (hours since midnight)",
    breaks = 4*(0:6)
) + geom_hline(
    "hline",
    yintercept = cosOut$estimates$mesor_cos, color = "red"
) + geom_vline(
    xintercept = cosOut$estimates$phi_cos, color = "orange"
) + geom_segment(
    aes(
        yend = cosOut$estimates$mesor_cos - 0.5*cosOut$estimates$amplitude_cos,
        y = cosOut$estimates$mesor_cos + 0.5*cosOut$estimates$amplitude_cos,
        x = cosOut$estimates$phi_cos + 12, 
        yend = cosOut$estimates$phi_cos + 12
    ), color = "green", arrow = arrow(ends = "both")
) + geom_label(
    data = tibble(
        name = c("mesor_cos", "amplitude_cos", "phi_cos"),
        x = c(
            20,
            cosOut$estimates$phi_cos + 12, 
            cosOut$estimates$phi_cos
        ),
        y = c(
            cosOut$estimates$mesor_cos, 
            cosOut$estimates$mesor_cos + 0.5*cosOut$estimates$amplitude_cos + 3,
            cosOut$estimates$mesor_cos + 0.5*cosOut$estimates$amplitude_cos + 10
        )
    ),
    aes(
        x = x,
        y = y,
        label = name
    )
) + ggtitle(
    "Basic Cosinor Estimates from extendedCosinorModel"
)
```

### Comparing Basic and Extended Cosinor Model Estimates

The figure above shows that the extended cosinor model (dashed line) appears to fit the observed data
better than the basic cosinor model (solid line).
The estimates of the fitted amplitude, acrotime, and mesor can be compared between both models.

```{r}
cosOut$estimates %>% transmute(
    amplitude_ext, amplitude_cos, 
    acrotime_ext, phi_cos,
    mesor_ext, mesor_cos
)
```

The pseudo F-statistic indicates the degree to which the extended cosinor model 
fits the data better than the basic cosinor model. 
`R2_ext` tells us that `r round(100*cosOut$estimates$R2_ext,1)`% of the variation 
in the epoch level data can be predicted by the extended cosinor fit.
`R2_cos` tells us that `r round(100*cosOut$estimates$R2_cos,1)`% of the variation 
in the epoch level data can be predicted by the basic cosinor fit.

```{r}
cosOut$estimates %>% select(
    F_pseudo_ext, R2_ext, R2_cos
)
```
